
# Dose 

There are three basic and interwoven principles of toxicology: 

1) dose-response 
2) `hazard x exposure = risk`
3) individual sensitivity. 

While these principles may form much of the foundation of toxicology, 
when it comes to any specific substance there is
likely to be controversy. 
Disagreement may arise on the relative importance of any one of
these principles while trying to evaluate implications for public health. 
Exploring these principles is an essential first step before examining their application to any specific substance. 



 
## Dose Response

In toxicology, the focus is usually on adverse reaction or response, 
but it is equally useful to consider a full range of responses from 
desirable to undesirable. Experience teaches us how to moderate the 
dose to achieve a desired result or avoid an undesirable effect.


Paracelsus stated correctly that the “... dose differentiates a poison 
from a remedy”, it is the individual that must constantly be aware of the dose 
and their particular response.


Dose is the amount of exposure to an agent, a quantitative measure of the 
exposure related to the subject or individual. For a chemical agent or drug 
the dose is the amount of the material in relation to body weight. 
Typically, the amount of material is measured in grams or thousandths of a gram
(milligrams, mg) and body weight is measured in kilograms (kg), 
equal to one thousand grams. The dose is the amount of material consumed divided by
body weight or mg/kg.

Calculate the dose
```
amt_mg = amount of material consumed (mg)
wt_kg = body weight in (kg)

oral_dose = amt_mg / wt_kg
```

### Coffee Caffeine Example

There are approximately 100 mg of caffeine in a cup of coffee. The actual amount
of caffeine in a cup of coffee depends on the coffee bean, how the coffee was prepared
and the size of the cup. Let us use an adult weighing 155 lbs (about 70 kg) 
consuming this one-cup of coffee.

```{r}
# kg = lbs / 2.205

amt_mg = 100
wt_kg =  70.30

oral_dose = function(amt_mg, wt_kg){
  dose = amt_mg / wt_kg
  glue::glue("Oral Dose = {round(dose,2)} mg/kg")
}

oral_dose(amt_mg, wt_kg)
```
This adult would receive a dose of 1.42 mg/kg of caffeine. 



### Difficulties with dose calculations

For our example of coffee caffeine:

- determining the exact amount of exposure to the agent
- amount of caffeine in a cup of coffee varies 
- the coffee bean type 
- brewing method
- size of the cup


Analytical chemists can accurately determine the amount of a specific 
agent in a material. 
Some foods, such as table salt or sugar, are relatively pure and the 
dose easily calculated by weighing the material. Package labeling usually 
indicates how many milligrams of the drug each pill contains, so the dose can 
be calculated.



Calculating the dose following workplace or environmental exposure can be far more
difficult. If the agent is in the air, then calculation of the dose must 
consider not only the concentration in the air but also the duration of 
the exposure, rate of breathing and body weight. The amount of air inhaled over a
period of time is estimated from laboratory data. Given this information, it is
possible to estimate the dose according to the following formula:

Airborne agent
```
inhal_dose    = Inhalation dose (mg/kg)
air_concent   = Air concentration of agent (mg/ml)
vol_air       = volume of air inhaled per hour (ml/hr)
dur_expos     = duration of exposure (hr)
b_wt          = body weight (kg)

inhal_dose = air_concent * vol_air * dur_expos / b_wt
```


For non-chemical exposures, other variables and different units of measurement are
required. Workplace and environmental exposures are often repeated and ongoing 
over an extended period of time. The health effects of repeated long-term exposures 
can be very different from one short-term exposure.



### Acute vs Chronic

For many drugs, we are looking for the immediate or acute response 
following exposure. We consume common painkillers with the desire to 
quickly stop our headache. Long-term repeated use, however, can have 
undesirable effects on the stomach or liver.

- `Acute exposure` -  is a single or very limited number of exposures over a short period of time. 

- `Chronic exposure` is repeated exposure over a long period of time. 


There is often a range of responses associated with any particular agent. 

- The response that occurs will vary with the dose
- the duration of exposure
- the individual

The acute response to a single dose is often the easiest to characterize, 
but the response to multiple exposures over a long period of time may be 
the most important. For a long time, it was thought that the only serious
complication from childhood lead exposure was death resulting from high 
exposure. Subsequent research demonstrated that even small amounts of lead 
exposure during childhood could result in brain damage that lasts a lifetime.
Determining what responses are most important is a central aspect of many 
debates in toxicology.



## Demonstrating Dose Response

In general, it is true that for any individual, the greater the dose the 
greater the response.  It is important to know if a substance distributes 
into body water because we made up of approximately 75% water.
An individual’s response to the caffeine varies with the dose and 
corresponding amount of circulating caffeine. Body size is a critical factor 
in determining dose and any subsequent response.


```{r,warning=FALSE, echo=F, collapse=TRUE, message=FALSE}
library(tidyverse)

wine = data.frame(
  difficulty_walking = c(0,0,10,25,40,60,77,90,100),
  glasses_of_wine = seq(0,8)
)

wine %>% 
  ggplot( aes(x= glasses_of_wine, y= difficulty_walking)) +
  geom_line() +
  geom_point( col='orange', size=3)+
  scale_x_continuous(breaks = 0:8)+
  labs(title= 'Dose Response for Drinking Wine',
       x= 'Glasses of Wine',
       y='% Difficulty Walking')+
  ggdark::dark_mode()
```


This plot illustrates the critical relationship between dose and response. 
In this case, we define the response as difficulty in walking and the dose of
or exposure to alcohol as a glass of wine. 
To change from exposure to dose we would need to know the body weight and 
amount of alcohol in the glass of wine.


In toxicology, the dose at which one half or 50% of the population is affected 
is often calculated and used to compare the toxicity of different agents. 
In this example, 50% of the population is affected after exposure to 4.5 
glasses of wine. 

```{r, warning=FALSE, echo=F, collapse=TRUE, message=FALSE}
library(paletteer)

wine2 = data.frame(
  response = c(0,2,10,15,45,70,77,80,95,97,100),
  dose = seq(0,100, 10)
)

wine2 %>% 
  ggplot( aes(x= dose, y= response, col= response))+
  geom_line(show.legend = F)+
  geom_point(size= 3, show.legend = F)+
  ggdark::dark_mode()+
  scale_x_continuous(breaks = seq(0,100,10) )+
  labs(x="Dose (mg/kg)", y="% Response", title = "An Idealized Dose/Response")+
  scale_color_paletteer_c(`"ggthemes::Orange-Blue-White Diverging"`, direction = -1)

```
For doses less than 30 mg/kg have no effect and doses above 50 mg/kg effect half 
the people. The “S”’-shaped idealized dose / response plot, which is typical
of most types of exposure. This “S” shaped cure illustrates that at low doses 
there is little or no response while at high doses all individuals respond 
or demonstrate the effect. In this situation, 50% of the subjects respond at a 
dose of 42 mg/kg, while 99% of the subjects responded at 90 mg/kg. 


It is important to emphasize that if we repeat this experiment the results 
would be slightly different. Each individual varies from one time to the next 
and there is even greater variability between individuals.
Variability is a consistent theme in biology, complicating data analysis 
and interpretation of results. These variations lead to the need for 
statistical evaluation of data.




## Hazard and Risk

The biological effects of an agent often span a broad range from 
beneficial to harmful, depending on the dose and individual sensitivity. 
The scientific discipline of toxicology developed in an effort to understand 
and characterize the potentially harmful or hazardous properties of an agent. 

Risk is the probability of:

- injury
- disease
- loss of function
- death for an individual or population exposed to a hazardous substance or situation. 

An agent or situation that can produce or cause a harmful or adverse effect is a Hazard. 


`Hazard` is an intrinsic property of a substance and any particular substance may
have a range of hazards associated with it depending upon specific conditions or
circumstances. 


On a daily basis, we routinely confront a range of potentially hazardous
agents, cooking, cleaning, driving, etc. We use these potentially hazardous 
agents but are careful to avoid conditions that will result in the 
expression of their hazardous properties.Hazard and risk are linked by exposure. 
Reducing the hazard, the exposure, or both can lower risk. If there is no exposure, 
then there is no risk or possibility of harm.


### DDT Hazard

In the past, the hazard associated with any particular substance was related to immediate
or obvious harm. As our knowledge and experience increase, so too does our appreciation
of an agent’s ability to produce unexpected consequences or harm. Take for example
DDT, a very powerful pesticide useful in the eradication of mosquitoes. As Rachel
Carson so eloquently pointed out, DDT devastated bird populations not directly but
indirectly by thinning eggshells to such an extent that the eggshells failed. This resulted
in a devastating decline in bird populations, particularly for birds consuming animals.
Still later we learned that DDT was a very persistent chemical and highly soluble in fat.
DDT thus accumulated up the food chain and in this case birds at the top of the food
chain were most affected. 

Humans are also at the top of the food chain, and through a
variety of means, DDT ends up in the food supply and becomes stored in body fat. When
women are breast-feeding, fat and DDT are mobilized and become the food of nursing
infants, which represents a large dose to a small infant. We are still unsure of the
consequences of fetal exposure to DDT and its effects on the developing organism. Many
other fat-soluble chemicals, such as dioxin and PCBs are known to contaminate breast
milk. Lead is another example of a major public health disaster that occurred because the
consequences of low-level lead exposure to the developing nervous system were not
appreciated.



## Lethal Dose

Historically, potentially toxic agents have been ranked by their lethality, 
or the amount of material that causes death. In this measure, hazard is defined 
only as death, obviously only the grossest measure of an agent’s effect. 
Because of individual variability or susceptibility, a standardized measure is the 
dose (in units of mg/kg) that produces death in half of the subjects, a 50% response.

This is called an LD50 or lethal dose for 50% of the population. 
The LD50 is one measure of the toxicity of a substance, its capacity for
causing illness or death. The LD50 is usually determined on populations of test animals
such as rats and mice. Determination of an LD50 is based on a single acute exposure to
an agent and the single response of death. Although the LD50 can be useful in comparing
the gross hazards of agents, it is not necessarily relevant to a response produced 
by low-level chronic exposure.


Problem with LD50


LD50s are misleading if used as the only characterization of the toxicity of a substance.

- lead effects on the developing nervous system even at very low levels of exposure. 
- Aspirin is a commonly used over-the-counter medicine
- DDT pesticide (banned) toxic effects and persistence in the environment. 

Yet they have similar LD50s.

lists the LD50s of a variety of common agents. Since the LD50 is the amount
of material required to produce death, a higher LD50 implies a lower toxicity and vice
versa.

Approximate Acute LD50s of Some Common Chemical Agents

| Agent                             | LD-50 (mg/kg)        |
|-----------------------------------|----------------------|
| Ethyl alcohol                     | 10,000               |
| Salt (sodium chloride)            | 4,000                |
| Iron (Ferrous sulfate)            | 1,500                |
| Morphine                          | 900                  |
| Mothballs (paradichlorobenzene)   | 500                  |
| Aspirin                           | 250                  |
| DDT                               | 250                  |
| Cyanide                           | 10                   |
| Nicotine                          | 1                    |
| Tetrodotoxin (from fish)          | 0.01                 |
| Dioxin (TCDD)                     | 0.001 (some species) |
| Botulinum Toxin                   | 0.00001              |


Fortunately, the LD50 is no longer recognized as an adequate or even particularly useful
assessment of an agent’s ability to cause harm. Toxicologists have developed a wide
array of tests to determine if an agent can produce an adverse effect. 
A variety of tests are performed to evaluate the potential harmful effects 
across all organ systems.






## Routes of Exposure and Absorption

Exposure, like many of the terms in toxicology, has several difference aspects, 
the most important of which are:

1) route of exposure 
2) frequency of exposure
3) duration of exposure 

Exposure is also affected by absorption. 
Even though we may come in contact with an agent, 
if little is taken up into the body (or absorbed), there is little effect.
What is not absorbed (and even some of what is absorbed) may be excreted from the body 
by various routes, including the urine, feces, and sweat or through exhalation. 
Excretion reduces the effect because it lowers the amount of toxicant in the body, 
thus reducing exposure to sensitive organs.

There are two types of exposure that deserve special attention: 

- fetal exposure during pregnancy 
- exposure of the brain (blood-brain barrier)


In real life, however, we are often exposed to a mixture of chemical agents. 
Multiple agents may interact and effect absorption or how the body reacts 
to the chemical. The body has a very sophisticated system to metabolize and 
eliminate chemicals form the body; this system plays an important role in 
protecting us from hazardous substances.




There are four main routes of exposure: 

1) skin (or dermal) exposure
2) lung (inhalation) exposure 
3) oral (gastrointestinal) exposure
4) injection exposure, delivery of drugs or medication that cannot be taken orally


Injections can take several forms. 
An injection directly into a blood vessel bypasses most of the absorption barriers 
and the drug will have almost full and immediate access to the most organs of the body. 

- intramuscular (IM) medications which slows absorption as the drug is slowly taken up by the blood supplying the muscle. 

- subcutaneous (SC) injections can be made just under the skin, commonly used for allergy testing or tuberculin (TB) tests




### Skin Exposure Route

Skin is the largest organ of the body and does an amazing job of protecting us 
from most agents. However, the skin is an important route of exposure to some 
agents and also a site of highly adverse reactions. 

In many cases, the skin is an excellent barrier to chemical agents, 
but some solvents can readily penetrate the skin.
The advantage of a skin patch in that the drug will be absorbed a 
constant slow rate, thus keeping the drug blood levels relatively constant. 



### Lung Exposure Route

Inhalation is an excellent route of exposure to many agents, including the oxygen
essential for life. The lungs are very rich in blood to facilitate the absorption of oxygen
and thus allow the rapid absorption of other agents directly into the bloodstream, quickly
producing an effect. 

Carbon monoxide is a potentially lethal gas that can be generated in
the home by poorly ventilated heaters, faulty furnaces, or a car idling in an attached
garage. Carbon monoxide is readily taken up by the blood cells by the same mechanism
as oxygen. In fact, carbon monoxide binds to the hemoglobin in the blood cells better
than oxygen, so exposure can cause serious injury and even death through lack of oxygen
intake. 


### Ingestion Exposure Route

Ingestion of substances orally allows absorption from the stomach and intestines. 
This is a critical route of exposure for many agents, from essential carbohydrates,
proteins, and vitamins, to unwanted pesticides and lead. All that is ingested is not
necessarily absorbed, and absorption can be dependent on age.

The increased absorption of certain agents at different times of life is related to the
body’s demand for important elements. Alcohol and caffeine are readily absorbed by 
the stomach, making for two of the most popular drugs in our culture. Oral exposure 
also occurs through our food and drinking water, so it is imperative to have unpolluted
water and a safe food supply.



### Frequency Exposure

Frequency can refer not only to the number of times the exposure occurred, 
but also to the time between exposures. Frequent exposure of a short duration 
results in rapidly elevated blood levels of any agent (assuming it’s absorbed).

- drinking four beers within 15 minutes vs drinking four beers in four days.

It takes approximately 30 minutes to absorb the caffeine from a cup of 
coffee and reach your peak blood caffeine levels. The harmful or toxic effects of 
an agent are often dependent on the frequency of exposure and the time between exposures.



### Duration of Exposure

In toxicology, duration is usually divided into three periods: 

1) `acute exposure` (usually just one or two exposures of short duration) 
2) `sub-chronic exposure` (multiple exposures over many days or perhaps months)
3) `chronic exposure` (long-term or even lifetime exposure)

The terms acute and chronic are also used to characterize the time delay between 
exposure and the onset of symptoms.

Acute effects are those noticed directly following exposure and 
are usually easily related to the agent.

Chronic or long-term effects of an agent may occur years later and are 
often very difficult to attribute to a particular cause.






## Metabolism, Distribution & Excretion

Metabolism refers to an organism’s ability to change a substance
into different chemical parts or metabolites that are usually less toxic. 
In toxicology, metabolism refers to the body’s ability to reduce an agent into
parts that are either less harmful or more readily excreted, a process 
called _detoxification_. The most common route of excretion is through urine, although 
some agents can be excreted in the feces, sweat or even the breath. 


Distribution refers to where an agent goes in the body. Some agents such as pesticides
and PCBs accumulate in the fat. Metabolism, distribution, and excretion are linked 
aspects that are essential to predicting the adverse effects of an agent and 
thus determining the risk of exposure to it.


The primary organ for detoxification is the liver. The liver has a variety of 
specialized cells that produce enzymes to aid in the metabolism of toxic agents. 
These enzymes can break down toxic agents into smaller elements, making them less toxic. 
In some cases, the compounds are changed so that they are more easily filtered 
by the kidney and excreted in the urine.

Another important mechanism of detoxification is the attachment or binding of 
another compound to a toxic chemical to make it easier for the kidney to filter 
the compound out of the blood and excrete in the urine. 

- Chelators bind metals so that they are more readily excreted in the urine. The problem with chelators is that they are non-specific and bind useful
agents such as calcium.

























